ODGI: understanding pangenome graphs

### 2021-10-18 manuscript/main.tex
Results: minor edits and comments

%ODGI is designed to build and modify paths in parallel, keeping its representation in memory efficient.
The ODGI toolkit builds on existing approaches to efficiently store and manipulate variation graphs \citep{Garrison:2018}.
%It implements the \textsc{HandleGraph} model \cite{Eizenga_2020_BX}.
Like other efficient libaries presenting the \textsc{HandleGraph} model \citep{Eizenga_2020_BX}, its implementation rests on three key properties which hold for most pangenome variation graphs:

\begin{enumerate}
\item They are relatively sparse, with low average node degree.
\item They can be sorted so that most edges go between nodes that are close together in the sort order.
\item Paths are locally similar to each other.
%We incorporate concepts first introduced in the dynamic version of the Graph BWT \citep{Siren:2020} (GBWT), a scalable implementation of the graph extension of the positional Burrows-Wheeler transform \citep{Durbin_2014}.
\end{enumerate}

These properties have already been used to build efficient dynamic variation graph data structures \citep{Siren:2020,Eizenga_2020_BX}.
Sparsity (1) allows us to encode edges $E$ using adjacency lists rather than matricies or hash tables.
The local linear structure of the graph (2) lets us assign node identifiers that increase along the linear components of the graph, which supports a compact storage of edges and path steps as relativistic (usually small) differences rather than absolute (always large) integer identifiers.
Path similarity (3) allows us to write local compressors that reduce the storage cost of collections of path steps.

%Here, we build on this work and focus on issues that have arisen during our recent work on the Human Pangenome.
Here, we improve on these prior efforts, focusing on issues that arose during our work with high-quality assemblies that cover almost all parts of the genome \citep{Logsdon_2021,Nurk_2021}.
%A key design objective, motivated by experience working with graphs built from high quality eukaryotic genome assemblies, is to support graphs with very high path depth.
We find that it is necessary to support graphs that have regions of very high path depth.
Such motifs can occur in collapsed repeat structures generated by ambiguous sequence homology relationships in repeats found in the centromeres and other segmental duplications.
If we cannot process such regions, we have only two options---we can remove these regions, or we can leave them unaligned.
Neither of these solutions allows us to investigate their biological features.
To seamlessy represent such difficult regions, we follow the approach of the dynamic Graph BWT (GBWT) \citep{Siren:2020}, and built a node-centric, dynamic, compressed model of the paths.
This design supports node-local modification and update of the graph, which lets us build and transform the model in parallel.

We store the graph in a vector of node structures, each of which presents a node-local view of the graph sequence, topology, and path layout.
Expressed in terms of the variation graph $V$, ODGI's core $Node$ structure includes a decoder that maps the neighbors of each node to a dense range of integers.
For a given node $i$ and neighbor $j$, the decoder itself does not store the $id$s of $j$, but rather a compact representation of the relative difference between the node ids: $\delta = Node_i.id - Node_j.id$.
This keeps even the size of the encoding small, per common variation graph property (2).
We define the edges and path steps traversing the node in terms of this alphabet of $\delta$'s.
The structures in algorithm \ref{alg:structs} describes our encoding.

@@

To allow edit operations in parallel, each node structure includes a byte-width mutex $lock$.
All edit operations on the graph must touch at most two $Node$ structs at a time.
(Both our edge and path step representations are doubly-linked.)
To avoid deadlocks, we acquire the node locks in ascending $Node.id$ order and release them in desceding order.
In addition to node-local features of the graph, we must maintain some global information.
Specifically, we record the start and end of paths, as well as a name to path id mapping in lock-free hashtables.
The use of lock-free hashtables lets us avoid a global lock when looking up path or graph metadata, which would quickly become a bottleneck during parallel operations on the graph.
By avoiding global locks, we implement many of the operations in ODGI using maximum parallelism available.
This ``atomic'' approach to edit operations on the graph is key to enabling our method to scale to the largest genome graphs that we can currently build.

@@

ODGI provides a set of first-order interrogative and manipulative operations on pangenome graphs.
We have established these tools to support our exploration of graphs built from hundreds of large eukaryotic genomes.
They are practical and able to work with high levels of graph complexity, even with regions of very high path depth (10$^5$ to 10$^6$-fold coverage).
To further our understanding of these graphs, we have developed a set of methods that let us inspect and manipulate them.

@@

To provide a universal coordinate space for the graph, we project its nodes into a low-dimensional space (in practice, 1 or 2 dimensions) by learning the projection via a HOGWILD!, path-guided stochastic gradient descent (PG-SGD) model.
This approach updates node relative positions to best-match their distance in the paths running through the graph.
Based on this projection, we can trivially generate 1D and 2D visualizations.
We can also detect regions whose layout is distorted, and which thus may represent structural variation or assembly error.
ODGI can project vector and matrix representations of the graph relative to these learned coordinate spaces.

@@

We begin with a set of HandleGraph-based algorithms first establish in the VG toolkit.
These include algorithms for graph traversal, partition-finding, $k$-mer and character enumeration, sorting, and pruning.
However, ODGI is more focused than VG.
Unlike VG, we provide no methods to construct graphs, map reads to the graph, or derive variants.

>>>>

%ODGI is designed to build and modify paths in parallel, keeping its representation in memory efficient.
The ODGI toolkit builds on existing approaches to efficiently store and manipulate variation graphs \citep{Garrison:2018}.
%It implements the \textsc{HandleGraph} model \cite{Eizenga_2020_BX}.
Similar to other efficient libaries presenting the \textsc{HandleGraph} model \citep{Eizenga_2020_BX}, ODGI's implementation rests on three key properties which hold for most pangenome variation graphs:

\begin{enumerate}
\item They are relatively sparse, with low average node degree.
\item They can be sorted so that most edges go between nodes that are close together in the sort order.
\item They contain paths are locally similar to each other.
%We incorporate concepts first introduced in the dynamic version of the Graph BWT \citep{Siren:2020} (GBWT), a scalable implementation of the graph extension of the positional Burrows-Wheeler transform \citep{Durbin_2014}.
\end{enumerate}

These properties are used to build efficient dynamic variation graph data structures \citep{Siren:2020,Eizenga_2020_BX}.
Sparsity (1) allows us to encode edges $E$ using adjacency lists rather than matrices or hash tables.
The local linear structure of the graph (2) lets us assign node identifiers that increase along the linear components of the graph, which supports a compact storage of edges and path steps as relativistic (usually small) differences rather than absolute (always large) integer identifiers.
Path similarity (3) allows us to write local compressors that reduce the storage cost of collections of path steps.
\FIXME{without a figure we may lose the reader here a little}

%Here, we build on this work and focus on issues that have arisen during our recent work on the Human Pangenome.
ODGI improves on prior efforts, based on issues that arose during our work with high-quality assemblies that cover almost all parts of the genome \citep{Logsdon_2021,Nurk_2021}.
%A key design objective, motivated by experience working with graphs built from high quality eukaryotic genome assemblies, is to support graphs with very high path depth.
For example, we find that it is necessary to support graphs that have regions of very high path depth.
Such motifs can occur in collapsed repeat structures generated by ambiguous sequence homology relationships in repeats found in the centromeres and other segmental duplications.
If we cannot process such regions, we have only two options---we can remove these regions, or we can leave them unaligned.
Neither of these solutions allows us to investigate their biological features.
To seamlessy represent such difficult regions, we follow \FIXME{or introduce?} the approach of the dynamic Graph BWT (GBWT) \citep{Siren:2020}, and built a node-centric, dynamic, compressed model of the paths.
\FIXME{So this has never been done before?}
This design supports node-local modification and update of the graph, which lets us build and transform the model in parallel.
\FIXME{Maybe plug ODGI for we?}

We store the graph in a vector of node structures, each of which presents a node-local view of the graph sequence, topology, and path layout.
\FIXME{Some repetition here?}
Expressed in terms of the variation graph $V$, ODGI's core $Node$ structure includes a decoder that maps the neighbors of each node to a dense range of integers.
For a given node $i$ and neighbor $j$, the decoder itself does not store the $id$s of $j$, but rather a compact representation of the relative difference between the node ids: $\delta = Node_i.id - Node_j.id$.
This keeps even the size of the encoding small, per common variation graph property (2).
We define the edges and path steps traversing the node in terms of this alphabet of $\delta$'s.
The structures in algorithm \ref{alg:structs} describes our encoding.

@@

To allow edit operations in parallel, each node structure includes a byte-width mutex $lock$.
All edit operations on the graph must touch at most two $Node$ structs at a time.
(Both our edge and path step representations are doubly-linked.)
To avoid deadlocks, we acquire the node locks in ascending $Node.id$ order and release them in desceding order.
In addition to node-local features of the graph, we must maintain some global information.
Specifically, we record the start and end of paths, as well as a name to path id mapping in lock-free hashtables.
The use of lock-free hashtables lets us avoid a global lock when looking up path or graph metadata, which would quickly become a bottleneck during parallel operations on the graph.
By avoiding global locks, we implement many of the operations in ODGI using maximum parallelism available.
This ``atomic'' approach to edit operations on the graph is key to enabling our method to scale to the largest genome graphs that we can currently build.
\FIXME{This is great!}

@@

ODGI provides a set of first-order interrogative and manipulative operations on pangenome graphs.
We have established these tools to support our exploration of graphs built from hundreds of large eukaryotic genomes.
\FIXME{why not just say human, mouse, \ldots?}
They are practical and able to work with high levels of graph complexity, even with regions of very high path depth (10$^5$ to 10$^6$-fold coverage).
To further our understanding of these graphs, we have developed a set of methods that let us inspect and manipulate them.

@@

To provide a universal coordinate space for the graph, we project its nodes into a low-dimensional space (in practice, 1 or 2 dimensions) by learning the projection via a HOGWILD!, path-guided stochastic gradient descent (PG-SGD) model.
\FIXME{Cite or explain?!}
This approach updates node relative positions to best-match their distance in the paths running through the graph.
Based on this projection, we can trivially generate 1D and 2D visualizations.
We can also detect regions whose layout is distorted, and which thus may represent structural variation or assembly error.
ODGI can project vector and matrix representations of the graph relative to these learned coordinate spaces.

@@

We begin with a set of HandleGraph-based algorithms first establish in the VG toolkit.
These include algorithms for graph traversal, partition-finding, $k$-mer and character enumeration, sorting, and pruning.
However, ODGI is more focused than VG.
Unlike VG, we provide no methods to construct graphs, map reads to the graph, or derive variants.
\FIXME{Above switches from ODGI to WE --- why not hold on to the former?}
